A

### 2015-10-18 paper-draft/paper.tex
add more images and update paper

@@

\section{Related work}

\section{Compact Finite Difference Schemes}

@@

\begin{equation} \label{eqn:compact-tridiagonal-system}
\begin{bmatrix}
     1 &  \alpha_1  \\
     \alpha   &  1   &  \alpha \\
         &  \alpha   &  1  &  \alpha  \\
         &      &  \alpha  &  1  &  \alpha  \\
         &      &     &     &  \ddots \\
         &      &     &     &     &  \ddots  \\
         &      &     &     &     &  \alpha_n &  1
\end{bmatrix}
\begin{bmatrix}
    f^{\prime}_1 \\
    f^{\prime}_2 \\
    f^{\prime}_3 \\
    \vdots \\
    \vdots \\
    f^{\prime}_{n-1} \\
    f^{\prime}_n
 \end{bmatrix}
=
\begin{bmatrix}
   d_1 \\
   d_2 \\
   d_3 \\
   \vdots \\
   \vdots \\
   d_{n-1} \\
   d_{n}
\end{bmatrix}
\end{equation}

@@

\section{Evaluation of compact finite differences}

@@

\section{Cyclic reduction}

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[height=200pt]{img/cyclic-reduction.eps}
\end{center}
\caption{Cyclic reduction -
the values that are solved for at each step of backward reduction
are highlighted}
\label{fig:cyclic-reduction}
\end{figure}

\begin{align} \label{eqn:forward-reduction}
& a^{\prime}_i = -a_{i-1}k_1 & \\
& b^{\prime}_i = b_i - c_{i-1}k_1 - a_{i+1}k_2 & \\
& c^{\prime}_i = -c_{i+1}k_2 & \\
& d^{\prime}_i = d_i - d_{i-1}k_1  - d_{i+1}k_2 &
\end{align}

\begin{align} \label{eqn:back-substitution}
x_i = \frac{d^{\prime}_i - a^{\prime}_ix_{i-1} - c^{\prime}_ix_{i+1}}{b^{\prime}_i}
\end{align}

At each step, for the last index $i=n$,
the forward reduction step is instead:
\begin{align} \label{eqn:forward-reduction-last}
    & a^{\prime}_n = -a_{n-1}k_1 & \\
    & b^{\prime}_n = b_n - c_{n-1}k_1 & \\
    & d^{\prime}_n = d_n - d_{n-1}k_1 &
\end{align}

While cyclic reduction performs far more work per step
compared to the Thomas algorithm,
it is able to solve the system in far fewer steps
on massively parallel architectures.
As pointed out by \cite{Zhang2010FTS},
$n$ parallel processors can solve a tridiagonal system
of size $n$ in $2log(n) - 1$ steps.
The Thomas algorithm, on the other hand,
requires $2n$ steps.

\section{Cyclic reduction for near-Toeplitz systems}

We note from
Equations \ref{eqn:forward-reduction}-\ref{eqn:back-substitution-last}
that each forward reduction step involves 12 operations.
As the LHS matrix is constant,
it is advantageous to pre-compute the coefficients appearing
in the forward reduction step.
For a general tridiagonal matrix,
this requires a substantial amount of storage
(2$n$ for each of the arrays $a$, $b$ and $c$):
$\frac{n}{2} + \frac{n}{4} + \frac{n}{8} + ... = n$
for the reduction steps,
and an additional $n$ for the original coefficients.
\section{Distributed compact-finite difference solution}

@@

\subsection{Evaluation of the right-hand side}

\subsection{Evaluation of the right-hand side}

\subsection{Solution of primary Near-Toeplitz tridiagonal systems}

\subsection{Solution of reduced system}

\subsection{Solution of secondary systems}

\section{Results}

>>>>

%----------------------------------------------------------------------%

@@

Compact finite difference schemes, developed by
Lele\cite{lele1992compact} 
are extremely popular in the computational fluid dynamics community
for their high order of accuracy at relatively small stencil widths.
These schemes require the solution of
linear systems that are
in general, \emph{banded}, and specifically,
for several popular schemes, \emph{tridiagonal} in nature.
Tridiagonal systems also appear in other numerical schemes
such as
Alternating Direction Implicit (ADI) methods
\cite{1955ADI}
and numerical solutions to one-dimensional differential equations.
In such CFD codes,
the evaluation of these tridiagonal systems constitute
a significant portion of the runtime.
Recently, there has been considerable interest in
the CFD community
in developing tridiagonal solvers
that exploit the massive parallelism afforded by
Graphics Processing Units (GPUs)
to speed up the computations
\cite{tutkun2012gpu}
\cite{esfahanian2014efficient}
\cite{GoSt11CR}.
The performance of tridiagonal solvers on the GPU is
dependent on several factors,
such as
choice of algorithm and the associated complexity,
thread utilization,
use of the available memory heirarchy,
and synchronization and control costs
\cite{Zhang2010FTS}.
The different approaches described in the literature
seek to make tradeoffs between these factors
to arrive at a solver that best fits the requirements.
But to our knowledge, the algorithms proposed so far
are applicable to general tridiagonal systems,
and make no attempt to take advantage of the known structure
of the tridiagonal matrix.
The tridiagonal systems that appear in several numerical schemes
are known to have a \emph{near-Toeplitz} structure.
We describe the implementation of a GPU solver that
makes use of this fact to
reduce the number of computations and
improve memory access.
The resulting solver is thus able to perform twice as fast as
the NVIDIA CUSPARSE equivalent
and $<$add a comparison to the CPU Thomas algorithm$>$.

This paper is organized as follows:

Section \ref{sec:preliminaries}
lists related works,
describes the cyclic reduction algorithm
and the specific issues in its implementation on GPUs.
Section \ref{sec:proposed-algorithm}
describes our proposed solver for near-Toeplitz tridiagonal systems.
Section \ref{sec:results-single-gpu}
provides an overview of the performance of our solver,
and compares it with other implementations.

%----------------------------------------------------------------------%
\section{Preliminaries} \label{sec:preliminaries}

The classical algorithm for solving tridiagonal systems is
the Thomas algorithm.
The algorithm itself exhibits no parallelism,
and while the GPU can still be exploited to solve
several independent tridiagonal systems,
the performance remains poor.
Zhang et al. \cite{Zhange2010FTS}
describe the implementation of three different algorithms:
cyclic reduction, parallel cyclic reduction, and
recursive doubling,
along with hybrid variants,
on GPUs.
These algorithms perform more work per step than
the Thomas algorithm,
but exhibit more fine-grained parallelism
that allows them to finish in
far fewer than the $2n$ steps required by the Thomas algorithm.
In the best case ($n$ parallel processors),
cyclic reduction requires 
$2log(n) - 1$ steps,
and parallel cyclic reduction requires
$log(n)$ steps
(while doing more work per step than cyclic reduction).
This feature makes them a natural choice for
implementation on GPUs.

\begin{figure}[h!]
\begin{center}
\includegraphics[height=200pt]{img/cyclic-reduction.eps}
\end{center}
\caption{Cyclic reduction - the highlighted elements
show the order in which values are solved for}
\label{fig:cyclic-reduction}
\end{figure}

\begin{align} \label{eqn:forward-reduction}
& a^{\prime}_i = -a_{i-1}k_1 & \\
& b^{\prime}_i = b_i - c_{i-1}k_1 - a_{i+1}k_2 & \\
& c^{\prime}_i = -c_{i+1}k_2 & \\
& d^{\prime}_i = d_i - d_{i-1}k_1  - d_{i+1}k_2 \
    \label{eqn:forward-reduction-rhs}&
\end{align}

\begin{align} \label{eqn:back-substitution}
x_i = \frac{d^{\prime}_i - a^{\prime}_ix_{i-1} - \
    c^{\prime}_ix_{i+1}}{b^{\prime}_i}
\end{align}

At each step, for the last index $i=n$,
the forward reduction step is instead:
\begin{align} \label{eqn:forward-reduction-last}
    & a^{\prime}_n = -a_{n-1}k_1 & \\
    & b^{\prime}_n = b_n - c_{n-1}k_1 & \\
    & d^{\prime}_n = d_n - d_{n-1}k_1&
\end{align}

%----------------------------------------------------------------------%
\section{Proposed algorithm} \label{sec:proposed-algorithm}

%----------------------------------------------------------------------%
\section{Results: single GPU} \label{sec:results-single-gpu}

%----------------------------------------------------------------------%

@@

\begin{equation} \label{eqn:compact-tridiagonal-system}
\begin{bmatrix}
     1 &  \alpha_1  \\
     \alpha   &  1   &  \alpha \\
         &  \alpha   &  1  &  \alpha  \\
         &      &  \alpha  &  1  &  \alpha  \\
         &      &     &     &  \ddots \\
         &      &     &     &     &  \ddots  \\
         &      &     &     &     &  \alpha_n &  1
\end{bmatrix}
\begin{bmatrix}
    f^{\prime}_1 \\
    f^{\prime}_2 \\
    f^{\prime}_3 
    \vdots \\
    \vdots \\
    f^{\prime}_{n-1} \\
    f^{\prime}_n
 \end{bmatrix}
=
\begin{bmatrix}
   d_1 \\
   d_2 \\
   d_3 \\
   \vdots \\
   \vdots \\
   d_{n-1} \\
   d_{n}
\end{bmatrix}
\end{equation}

@@

%----------------------------------------------------------------------%

@@

%----------------------------------------------------------------------%

@@

%----------------------------------------------------------------------%

%----------------------------------------------------------------------%

We note from
Equations \ref{eqn:forward-reduction}-\ref{eqn:forward-reduction-rhs}
that each forward reduction step involves 12 operations.
As the LHS matrix is constant,
it is advantageous to pre-compute the coefficients appearing
in the forward reduction step.
For a general tridiagonal matrix,
in addition to the $3n$ original coefficient arrays
$a$, $b$ and $c$,
this requires extra storage of 
$3(\frac{n}{2}+\frac{n}{4}+\frac{n}{8}+...)=3n$
coefficients.
This storage cost is greatly amortized
when solving multiple right-hand sides.

@@
